Coherent Hyper-Raman Four Wave Mixing Vibrational Spectroscopies

### 2024-07-08 paper.tex
wip
added more on abcs but need to do vibronic coupling still

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{dfg_hdfg_hr_wmel.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of difference frequency generation (DFG), hyper difference frequency generation (HDFG) and spontaneous hyper-Raman scattering. \cite{RN286, RN352}
		$\ket{g}$ is the ground state, $\ket{v}$ is an infrared active vibration and $\ket{m}, \ket{n}$ are virtual states.
		Solid and dashed horizontal lines indicate real and virtual states, whereas solid and dotted arrows indicate ket and bra side transitions, respectively. %todo: add Raman to this WMEL! dfg Raman hyperDFG hyperRaman 
		}
	\label{fig:comparisonwmel}
\end{figure}

Not all CMDS methods are used to dissect intramolecular couplings.\cite{Shen1987_CPL}
Non-parametric, hyper difference frequency generation (hyper-DFG) spectroscopy, similar to difference frequency generation (\autoref{fig:comparisonwmel}), is a four wave mixing method which, when detuned from  electronic resonances, does not probe intramolecular couplings. 
HDFG was the first infrared CMDS method to successfully discriminate against non-resonant background.\cite{RN351, RN352}
Hyper-DFG has also been referred to as singly vibrationally enhanced (SIVE) spectroscopy. \cite{RN351}
In hyper-DFG, an infrared pulse is resonant with a vibrational mode, and two other input pulses are used to induce a scattering process to provide output.
Hyper-DFG was documented long ago to have characteristics similar to spontaneous hyper-Raman scattering spectroscopy. \cite{RN352}
Hyper-Raman scattering is the two photon analogue to Raman scattering. \cite{Cyvin1965, Terhune1965}
Compared to Raman spectroscopy, hyper-Raman scattering cross sections are often small, making it a difficult technique to successfully implement.\cite{RN515, Kelley2010} 
The different selection rules of hyper-Raman scattering, however, make it a unique alternative to Raman scattering to understand vibrational spectra and vibronic coupling in molecular systems.
In 1998, Cho et al. proposed a parametric six wave mixing technique which was the coherent analogue of hyper-Raman scattering. \cite{Cho1998}
However, since most six wave mixing methods cascade into four wave mixing processes,\cite{RN243, Cho2000_Cascade} it is preferable to investigate coherent four wave mixing analogues to hyper-Raman scattering.

While hyper-DFG showed promise as an upconverted infrared spectroscopy method, it was quickly supplemented by a doubly vibrationally resonant method, doubly vibrationally enhanced (DOVE) spectroscopy, to investigate intra- and intermolecular vibrational coupling. \cite{RN345, RN101, Cho2000}
As a result, investigations of hyper-DFG stagnated; to date, there are less than a dozen studies which report on hyper-DFG. \cite{RN350, RN416, RN351, RN352, RN353, Chen1998, RN362, RN418, Bonn2024, McDonnell2024}
Recent reports highlight non-negligible hyper-DFG interference in DOVE experiments as well as the suitability of hyper-DFG pathways to interpret ultrafast relaxation dynamics in isotropic systems. \cite{Bonn2024, McDonnell2024}
These reports motivate deeper analysis on the properties of hyper-DFG spectroscopy.

@@

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{wmel1.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of the relevant HDFG pathways in the (a) $\vec{k}_4 = -\vec{k}_1 + \vec{k}_2 + \vec{k}_3$ and (b) $\vec{k}_4 = \vec{k}_1 + \vec{k}_2 + \vec{k}_3$ phasematching geometries. \cite{RN286, RN352}
	$\ket{g}$ is the ground state, $\ket{v}$ is an infrared active vibration and $\ket{m}, \ket{n}$ are virtual states.
	Solid and dashed horizontal lines indicate real and virtual states, whereas solid and dotted arrows indicate ket and bra side transitions, respectively. 
	These diagrams can be permuted to generate additional resonant WMEL diagrams.}
	\label{fig:sivewmel}
\end{figure}

@@

To make the connection between HDFG and hyper-Raman scattering, we investigate its gross selection rules.
By propagating density matrix elements in the steady state limit, the HDFG hyperpolarizability is \cite{RN119}
\begin{equation}\label{sivegamma}
		\gamma_{ijkl} =	- \sum_{m, n} \frac{1}{\varepsilon_0} \frac{1}{4D} \frac{1}{\hbar^3} \frac{\mu^{i}_{v n} \mu^{j}_{nm} \mu^{k}_{mg} \mu^{l}_{gv} }{\Delta_{nv} \Delta_{mv}\Delta_{gv}}  \rho_{gg}
\end{equation}
where: $\mu^{j}_{ab}$ is the $j^{th}$ element of $\mel{a}{\vec{\mu}}{b}$, $\Delta_{kl} = \omega_{kl} - \omega_{j} - i\Gamma_{kl}$, $\omega_j$ is the frequency of the j$^{th}$ input field, $\Gamma_{kl}$ is the dephasing of $\rho_{kl}$ and $\rho_{gg}$ is the ground state population.
For simplicity, we consider the results for the process diagrammed in \autoref{fig:sivewmel}a, but by multiplying by (-1) and taking a complex conjugate, which arise from the bra-side transition in \autoref{fig:sivewmel}a, results for the \autoref{fig:sivewmel}b  process are obtained.
Note that $\omega_j \rightarrow -\omega_j$ for a bra side transition.
D, the Maker-Terhune degeneracy factor, accounts for permutation symmetry in $\gamma_{ijkl}$ and is defined as $\frac{n!}{(n-m)!}$, where $n$ is the total number of input fields (3 for any four wave mixing experiment) and $m$ is the number of distinguishable input fields.\cite{RN134} 
For a HDFG experiment using two or three distinct input fields, as considered here, D = 6.
Note that $\ket{m}, \ket{n}$ are virtual states (\autoref{fig:sivewmel}).
By contracting over the virtual states to form the hyper-Raman hyperpolarizability $\beta$ (i.e., Placzek approximation), this expression can be simplified immensely.\cite{Long1970} 
Note, however, that if $\omega_2$, $\omega_3$, as depicted in \autoref{fig:sivewmel}, are nearly resonant with any state, the Placzek approximation fails. \cite{Placzek1934, Long1970}
Assuming the validity of the Placzek approximation, \autoref{sivegamma} is written as 
\begin{equation}\label{sivebeta}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{24 \hbar}\frac{\beta^{(ijk)}_{gv} \mu^{(l)}_{gv}}{\Delta_{gv}} \rho_{gg}
\end{equation}
Taylor expanding the dipole and first hyperpolarizability operators in terms of an n$^{\text{th}}$ normal mode coordinate $Q_n$ about equilibrium: \cite{Long1970}
\begin{subequations}
	\begin{equation}
		\mu = \mu_0 + \frac{\partial \mu}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
	\begin{equation}
		\beta = \beta_{0} + \frac{\partial \beta}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
\end{subequations}
Substituting into \autoref{sivebeta}, and knowing that $\mel{v}{Q_n}{g} = \sqrt{\frac{\hbar}{2m_n\omega_{vg}}}$ under the harmonic approximation,\cite{RN230} where $m_n$ is the mass of the $Q_n$ mode, gives the HDFG hyperpolarizability to $\order{Q_n}$ as \begin{equation}\label{SIVEselection}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{48 m_n \omega_{vg}}  \frac{1}{{\Delta_{gv}}} \ \frac{\partial \beta^{(ijk)}}{\partial Q_n} {\frac{\partial \mu^{(l)}}{\partial Q_n}}  \rho_{gg}
\end{equation}
Since this expression is non-zero in the harmonic oscillator limit, HDFG output is allowed for harmonic transitions. 
Similar to SFG, HDFG does not demand anharmonic ground state potential energy surfaces for spectral output. \cite{Shen94, Cho2000}
This selection rule is generally valid for any HDFG or hyper sum frequency generation (HSFG) process in the $-\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ and $\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ phasematching geometries when the Placzek approximation is valid.

\autoref{SIVEselection} shows any HDFG active mode must be both hyper-Raman and IR active.
However, since any IR active vibrational mode is hyper-Raman active,\cite{Cyvin1965, Andrews1978} any IR active mode is HDFG active.
This makes HDFG an optically upconverted infrared spectroscopy of isotropic samples, as its selection rules are identically those of IR spectroscopy, but its output is generally in the visible.
A useful attribute of the $\beta_{ijk}$ dependence in HDFG output is added polarization control. 
Similar to SHG and SFG experiments that probe different elements of the infrared transition dipole and Raman polarizability tensors,\cite{Heinz1982} probing different elements of $\beta_{ijk}$ in HDFG can greatly assist in understanding ultrafast dynamics in isotropic systems, as suggested by Seliya et al. \cite{Shen90, Bonn2024}
Polarization control in IR-pump-HDFG-probe, analogous to ultrafast transient absorption spectroscopy for vibrational modes, would be an interesting venue to assess the sensitivity of HDFG to different hyper-Raman tensor elements.

Previous spontaneous hyper-Raman work interpreted the origin of hyper-Raman signal in terms of Franck-Condon, Herzberg-Teller and Duschinsky type couplings.\cite{Ziegler1988} %todo: cite the other papers.
As such, we investigate \autoref{SIVEselection} using an Albrecht ABC type formalism which naturally incorporates these types of couplings.
To do so, we write the states in terms of a Born-Oppenheimer basis $\ket{a,b}$, where $\ket{a,b} = |a) \otimes \ket{b}$ for electronic states $\{|a)\}$ and vibrational states $\{\ket{b}\}$. \cite{BornOppenheimer, Albrecht1960}
Here, $\ket{g}, \ket{v}$ are vibrational states on the ground electronic state $|g)$ and $\ket{m} \rightarrow \ket{m,j}$, $\ket{n} \rightarrow \ket{n,k}$ are virtual states.
For consistency with previous reports, we use $\vec{R}$ to denote electric transition dipole moments ($\vec{\mu}$ is used for transitions between states on the ground electronic state.). \cite{Ziegler1974}
Following the approach which gave \autoref{sivegamma}, we find
\begin{equation}\label{drgamma_notaylor}
	\gamma_{ijkl} = -\frac{1}{\varepsilon_0} \frac{1}{24 \hbar^3} \sum_{m,n,j,k} \frac{
		R^{i}_{gv, nk} 
		R^{j}_{nk,mj} 
		R^{k}_{mj,gg} 
		R^{l}_{gv,gg} 
	}{\Delta_{g0,gg}
		\Delta_{nk, mj}
		\Delta_{mj, gg}
	}
\end{equation}
where $R^{i}_{ab,cd}$ is the i$^{th}$ element of $\mel{a,b}{\vec{R}}{c,d}$.\cite{Ziegler1988}
Using our definition of vibronic states, we write $\vec{M}_{ab} = (a|\vec{M}|b)$ so that, for example,
$R^{i}_{gv,ev'} = \mel{v}{M^i_{ge}}{v'}$.\cite{Albrecht1960}
Using these ideas rewrites \autoref{drgamma_notaylor} as
\begin{equation}\label{drgamma_notaylor_1}
	\begin{split}
		\gamma_{ijkl} &= -\frac{1}{\varepsilon_0} \frac{1}{24 \hbar^3} \sum_{m,n,v'} \frac{
			\mel{v}{M^{i}_{ge}}{v'} 
			\mel{v'}{M^{j}_{em}}{n}
			\mel{n}{M^{k}_{mg} }{v}
			\mel{v}{M^{l}_{gg}}{0}
		}{\Delta_{g0,gv}
			\Delta_{ev', mn}
			\Delta_{mn, g0}	} \\
	\end{split}
\end{equation}
Chung and Ziegler showed some years ago that expanding ${\vec{M}_{ij}}$ to $\order{Q}$ yields $A, B, C$ coefficients similar to the Albrecht formalism of Raman scattering. \cite{Albrecht1961, Ziegler1988}
The $A$ term depends upon Franck-Condon $\order{Q^0}$ transitions, the $B$ term depends upon Herzberg-Teller ($\order{Q}$) transitions, and the $C$ term depends on Duschinsky ($\order{Q^2}$) transitions. 
Here, we ignore the C term as it depends on one and two photon forbidden transitions which have yet to be observed in spontaneous hyper-Raman scattering. \cite{Ziegler1988}
The full derivation of $A,B$ have been given elsewhere, so only their final forms will be given.

By compiling all the terms that scale with $\order{Q^0}$, i.e. $\vec{M}_{ij_0}$, $A$ is written as
\begin{equation}
	\begin{split}
		A_{ijk} &= \text{crap times Franck Condon}\\
	\end{split}
\end{equation}
where terms such as $\langle v | v' \rangle$ are Franck-Condon factors.

something about the A term. SHOW THAT IT VANISHES FOR SINGLY RES TERMS

It is important to note that in the case of singly resonant HDFG, $A_{ijk}$ vanishes, or only Herzberg-Teller contributions are present in SR-HDFG. \cite{Neddersen1989}

Similarly, the $B$ terms scale with $\order{Q}$, such that
\begin{equation}
	\begin{split}
		B_{ijk} &= \text{crap}\\
	\end{split}
\end{equation}
Then B term. 
B term has new sum over another normal mode Ql that must be included because when I expand in terms of Q there are 3N-6-1 normal modes I can talk to (3N-6-1, assuming nonlinear and that the -1 comes from assuming we end in v)
Blah Blah Blah
%todo: So clearly HDFG has heavy dependence on HT type couplings! 
%This likely explains the relative weakness of HDFG in solvents such as hexane. \cite{RN351}
%fix the equation so that the dipole is clearly isolated from the hyper Raman part. Write beta as A+B+C?
%todo: segue into vibronic coupling

While HDFG can provide useful information about the properties of molecular vibrations, it is immediately apparent that \autoref{fig:sivewmel} can inform on molecular electronic structure by making the hyper-Raman transition resonant with a vibronic state.
DR-HDFG (where a vibrational and vibronic state are resonantly coupled) can provide a tool for measuring vibronic coupling.
This method is analogous to doubly resonant sum frequency generation (DR-SFG). \cite{Shen94}
Many multi-resonant nonlinear spectroscopies have been developed to resolve vibronic coupling in molecular samples. \cite{Carlson1990, Gaynor2017, RN276}
DR-HDFG provides methods to investigate vibronic coupling with only two laser pulses, as long as the electronic state is accessible via two photon absorption from the ground state (\textit{vide infra}).
Two beam, parametric and non-parametric DR-HDFG processes are diagrammed in \autoref{fig:doubsive}.
%todo: insert discussion on ABC terms
%todo: write hamiltonian like Gaynor to make clear what we are doing (and put in a picture in Figure doubbleresspec)

To interpret HDFG selection rules, we use a slightly modified version of equation xyz to account for the added vibronic resonance. 
This change immediately makes the A term nonzero in the case of noncentrosymmetric systems.

@@

\subsection{Quantification}
Since the hyper-Raman process is stimulated by a two-photon process, a setup using only two lasers can perform a coherent four wave mixing hyper-Raman experiment by investigating $\pm \vec{k}_1 + 2\vec{k}_2$ processes, assuming a two photon absorption from a single pulse stimulates the hyper-Raman process.
The other commonly used two laser CMDS method is sum frequency generation, a $\chi^{(2)}$ technique, whose phasematching is dictated by $\vec{k}_1 + \vec{k}_2$.
The SFG output scales as $\chi^{(2)}_{IJK} \sim \langle \alpha_{ij} \mu_k \rangle$, where $\alpha_{ij}$ is the Raman polarizability tensor.
Since SFG is dependent upon transitions that transform like both rank-one and rank-two tensors, this enforces a microscopic constraint that SFG cannot yield output for centrosymmetric species under the electric dipole approximation.
This follows from the presence of an inversion center in centrosymmetric systems, which demands strict parity selection rules. \cite{RN230}
Macroscopically, inversion symmetry in the output polarization eliminates output from centrosymmetric species under the electric dipole approximation.\cite{RN227, RN132}
As a result, the output of SFG is significantly reduced relative to most third order spectroscopies because the output only depends on surface species.
However, since HDFG is dependent upon a hyper-Raman transition, it is useful to compare the relative output of the two processes to identify how HDFG output compares to SFG.

@@

With the selection rules of HDFG understood for vibrational spectroscopy, it becomes possible to extract quantitative information from its spectra.
Lineshape analysis is essential for extracting quantitative information from CMDS spectra.
Scanning across resonances create dispersive lineshapes, i.e, self-heterodyning, which inform on $\Re(\chi^{(3)})$ and $\Im(\chi^{(3)})$.\cite{Levenson1974_1, Levenson1974_2}
In the method introduced by Levenson and Bloembergen (Bloembergen Interferometry Experiment), an internal standard interferes with the resonant lineshape.
The internal standard is often chosen to be C$_6$H$_6$ or C$_6$D$_6$, due to the strong Raman activity and measured $\chi^{(3)}$ of its ring breathing mode ($\nu_2$ in Herzberg notation). \cite{Levenson1974_2, RN351, RN345}
%okay clearly this wont work lol we need something that is IR active... CaF2?%
Self-heterodyning of the internal standard signal measures $\chi^{(3)}$ and does not require measurement of absolute intensities. 
Resonant lineshapes are also complicated by amplitude level interference between the sample and its substrate and/or sample cell windows, which must be accounted for to obtain quantitatively correct $\chi^{(3)}$ values. \cite{RN362, RN418}

>>>>

\begin{figure}[!htbp]
	\centering
	\includegraphics[width=3.375in]{f1.png}
	\caption{Wave Mixing Energy Level (WMEL) diagrams of difference frequency generation (DFG), spontaneous Raman scattering, hyper difference frequency generation (HDFG) and spontaneous hyper-Raman scattering. \cite{RN286, RN352}
		$\ket{g}$ is the ground state, $\ket{v}$ is an infrared active vibration and $\ket{m}, \ket{n}$ are virtual states.
		Solid and dashed horizontal lines indicate real and virtual states, whereas solid and dotted arrows indicate ket and bra side transitions, respectively. %todo: add Raman to this WMEL! dfg Raman hyperDFG hyperRaman 
		}
	\label{fig:comparisonwmel}
\end{figure}

Not all CMDS methods are used to dissect intramolecular couplings.\cite{Shen1987_CPL}
Non-parametric, hyper difference frequency generation (hyper-DFG) spectroscopy, similar to difference frequency generation (\autoref{fig:comparisonwmel}), is a four wave mixing method which, when detuned from  electronic resonances, does not probe intramolecular couplings. 
HDFG was the first infrared CMDS method to successfully discriminate against non-resonant background.\cite{RN351, RN352}
Hyper-DFG has also been referred to as singly vibrationally enhanced (SIVE) spectroscopy. \cite{RN351}
In hyper-DFG, an infrared pulse is resonant with a vibrational mode, and two other input pulses are used to induce a scattering process to provide output.
Hyper-DFG was documented long ago to have characteristics similar to spontaneous hyper-Raman scattering spectroscopy. \cite{RN352}
Hyper-Raman scattering is the two photon analogue to Raman scattering. \cite{Cyvin1965, Terhune1965}
Compared to Raman spectroscopy, hyper-Raman scattering cross sections are often small, making it a difficult technique to successfully implement.\cite{RN515, Kelley2010} 
The different selection rules of hyper-Raman scattering, however, make it a unique alternative to Raman scattering to understand vibrational spectra and vibronic coupling in molecular systems.
Some years ago, Cho et al. proposed parametric six wave mixing techniques as the coherent analogue of hyper-Raman scattering. \cite{Cho1997, Cho1998}
However, since most six wave mixing methods cascade into four wave mixing processes,\cite{RN243, Cho2000_Cascade}, and their output would be dependent upon the square magnitude of the hyper-Raman scattering cross section, a small number,\cite{RN515} it is preferable to investigate coherent four wave mixing analogues to hyper-Raman scattering.

While hyper-DFG showed promise as an upconverted infrared spectroscopy method, it was quickly supplemented by a doubly vibrationally resonant method, doubly vibrationally enhanced (DOVE) spectroscopy, to investigate intra- and intermolecular vibrational coupling. \cite{RN345, RN101, Cho2000}
As a result, investigations of hyper-DFG stagnated; to date`Z, there are less than a dozen studies which report on hyper-DFG. \cite{RN350, RN416, RN351, RN352, RN353, Chen1998, RN362, RN418, Bonn2024, McDonnell2024}
Recent reports highlight non-negligible hyper-DFG interference in DOVE experiments as well as the suitability of hyper-DFG pathways to interpret ultrafast relaxation dynamics in isotropic systems. \cite{Bonn2024, McDonnell2024}
These reports motivate deeper analysis on the properties of hyper-DFG spectroscopy.

@@

@@

To make the connection between HDFG and hyper-Raman scattering, we investigate its gross selection rules.
By propagating density matrix elements in the steady state limit, the HDFG hyperpolarizability is \cite{RN119}
\begin{equation}\label{sivegamma}
		\gamma_{ijkl} =	- \sum_{m, n} \frac{1}{\varepsilon_0} \frac{1}{4D} \frac{1}{\hbar^3} \frac{\mu^{vn}_{i} \mu^{nm}_{j} \mu^{mg}_{k} \mu^{gv}_{l} }{\Delta_{nv} \Delta_{mv}\Delta_{gv}}  \rho_{gg}
\end{equation}
where: $\mu^{ab}_{j}$ is the $j^{th}$ element of $\mel{a}{\vec{\mu}}{b}$, $\Delta_{kl} = \omega_{kl} - \omega_{j} - i\Gamma_{kl}$, $\omega_j$ is the frequency of the j$^{th}$ input field, $\Gamma_{kl}$ is the dephasing of $\rho_{kl}$ and $\rho_{gg}$ is the ground state population.
D, the Maker-Terhune degeneracy factor, accounts for permutation symmetry in $\gamma_{ijkl}$.\cite{RN134} 
For a HDFG experiment using two or three distinct input fields, as considered here, D = 6.
Note that $\ket{m}, \ket{n}$ are virtual states (\autoref{fig:sivewmel}).
By contracting over the virtual states to form the hyper-Raman hyperpolarizability $\beta$, this expression can be simplified immensely.\cite{Long1970} 
Note, however, that if $\omega_2$, $\omega_3$, as depicted in \autoref{fig:comparisonwmel}, are nearly resonant with any state, the contraction is not allowed. \cite{Placzek1934, Long1970}
Assuming $\omega_2$, $\omega_3$ are significantly detuned from any local resonance, \autoref{sivegamma} is written as 
\begin{equation}\label{sivebeta}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{24 \hbar}\frac{\beta^{gv}_{ijk} \mu^{gv}_{l}}{\Delta_{gv}} \rho_{gg}
\end{equation}
Taylor expanding the dipole and first hyperpolarizability operators in terms of an n$^{\text{th}}$ normal mode coordinate $Q_n$ about equilibrium: \cite{Long1970}
\begin{subequations}
	\begin{equation}
		\mu = \mu_0 + \frac{\partial \mu}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
	\begin{equation}
		\beta = \beta_{0} + \frac{\partial \beta}{\partial Q_n} Q_n + \order{{Q_m Q_n}}
	\end{equation}
\end{subequations}
Substituting into \autoref{sivebeta} gives the HDFG hyperpolarizability to $\order{Q_n}$ as \begin{equation}\label{SIVEselection}
	\gamma_{ijkl} =	-\frac{1}{\varepsilon_0} \frac{1}{48 m_n \omega_{vg}}  \frac{1}{{\Delta_{gv}}} \ \frac{\partial \beta^{gv}_{ijk}}{\partial Q_n} {\frac{\partial \mu^{gv}_{l}}{\partial Q_n}}  \rho_{gg}
\end{equation}
where $\mel{v}{Q_n}{g} = \sqrt{\frac{\hbar}{2m_n\omega_{vg}}}$ under the harmonic approximation,\cite{RN230} and $m_n$ is the ``mass'' of the $Q_n$ mode. 
Since this expression is non-zero in the harmonic oscillator limit, HDFG output is allowed for harmonic transitions. 
Similar to SFG, HDFG does not demand anharmonic ground state potential energy surfaces for spectral output. \cite{Shen94, Cho2000}
This selection rule is generally valid for any HDFG or hyper sum frequency generation (HSFG) process in the $-\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ and $\vec{k}_1 + \vec{k}_2  + \vec{k}_3$ phasematching geometries when $\omega_2$ and $\omega_3$ are sufficiently detuned from any resonance.

So far, only Placzek approximation type arguments have been used to interpret HDFG output.
Previous spontaneous hyper-Raman work interpreted the origin of hyper-Raman signal in terms of Franck-Condon, Herzberg-Teller and Duschinsky type couplings.\cite{Ziegler1988} 
As such, we extend \autoref{SIVEselection} using an Albrecht ABC type formalism to account for these couplings.
The formalism will first be developed for singly resonant HDFG, but extension to doubly resonant HDFG is straightforward and will be discussed later.

To employ the Albrecht formalism, we write the states in terms of a Born-Oppenheimer basis $\ket{a,b}$, where $\ket{a,b} = |a) \otimes \ket{b}$ for electronic states $\{|a)\}$ and vibrational states $\{\ket{b}\}$. \cite{BornOppenheimer, Albrecht1960}
For simplicity, we choose $\ket{g}, \ket{v}$ to be vibrational states on the ground electronic state $|g)$ and $\ket{m} \rightarrow \ket{m,j}$, $\ket{n} \rightarrow \ket{n,k}$ are virtual states.
For consistency with previous reports, we use $\vec{R}$ to denote electric transition dipole moments ($\vec{\mu}$ is reserved for transitions between states on the ground electronic state). \cite{Ziegler1974}
Following the approach which gave \autoref{sivegamma}, we find
\begin{equation}\label{drgamma_notaylor}
	\gamma_{ijkl} = -\frac{1}{\varepsilon_0} \frac{1}{24 \hbar^3} \sum_{m,n,j,k} \frac{
		R_{i}^{gv, nk} 
		R_{j}^{nk,mj} 
		R_{k}^{mj,gg} 
		R_{l}^{gv,gg} 
	}{\Delta_{g0,gg}
		\Delta_{nk, mj}
		\Delta_{mj, gg}
	}
\end{equation}
where $R_{i}^{ab,cd}$ is the i$^{th}$ element of $\mel{a,b}{\vec{R}}{c,d}$.\cite{Ziegler1988}
Using our definition of vibronic states, we write $\vec{M}^{ab} = (a|\vec{M}|b)$ so that, for example,
$R_{i}^{gv,ev'} = \mel{v}{M_i^{ge}}{v'}$.\cite{Albrecht1960}
Using these ideas rewrites \autoref{drgamma_notaylor} as
\begin{equation}\label{drgamma_notaylor_1}
	\begin{split}
		\gamma_{ijkl} &= -\frac{1}{\varepsilon_0} \frac{1}{24 \hbar^3} \sum_{m,n,v'} \frac{
			\mel{v}{M_{i}^{ge}}{v'} 
			\mel{v'}{M_{j}^{em}}{n}
			\mel{n}{M_{k}^{mg} }{v}
			\mel{v}{M_{l}^{gg}}{0}
		}{\Delta_{g0,gv}
			\Delta_{ev', mn}
			\Delta_{mn, g0}	} \\
%	&= 	-\frac{1}{\varepsilon_0} \frac{1}{24 \hbar^3} \sum_{m,n,v'} \frac{
%		\mel{v}{M_{i}^{ge}}{v'} 
%		\mel{v'}{M_{j}^{em}}{n}
%		\mel{n}{M_{k}^{mg} }{v}
%		\mel{v}{\mu_{l}}{0}
%	}{\Delta_{g0,gv}
%		\Delta_{ev', mn}
%		\Delta_{mn, g0}	} \\
	\end{split}
\end{equation}
Chung and Ziegler showed some years ago that expanding $\vec{M}^{ij}$ to $\order{Q}$ as
\begin{equation}
\vec{M}^{ij} = \vec{M}^{ij}_0 + \sum_z \frac{\partial\vec{M}^{ij}}{\partial Q_z}
\end{equation} yields $A, B, C$ coefficients similar to those in the Albrecht formalism of Raman scattering. \cite{Albrecht1961, Ziegler1988}
The $A$ term depends upon Franck-Condon ($\order{Q^0}$) transitions, the $B$ term depends upon Herzberg-Teller ($\order{Q}$) transitions, and the $C$ term depends on Duschinsky ($\order{Q^2}$) transitions. 
Here, we ignore the C term as it depends on one and two photon forbidden transitions which have yet to be observed in spontaneous hyper-Raman scattering. \cite{Ziegler1988}
The full derivation of $A,B$ have been given elsewhere, so only their final forms will be given.
In the discussion that follows, we take $\hbar = e = 1$ for simplicity.

By compiling all the terms that scale with $\order{Q^0}$, i.e. $\vec{M}^{ij}_0$, $A$ is written as
\begin{equation}
	\begin{split}
		A_{ijk} = \sum_{m,n,j,k} M^{gn}_{0,i} 
		M^{nm}_{0,j} 
		M^{mg}_{0,k}
		 \langle v | k \rangle
		 \langle k | j \rangle
		 \langle j | g \rangle \\
	\end{split}
\end{equation}
and $B$ is written as a sum of two terms, $B_1, B_2$, as
\begin{subequations}
	\begin{equation}
		\begin{split}
			B_{1_{ijk}} &= \sum_{m,n,j,k,z} M^{gn}_{0,i} \langle v | k \rangle \lbrack
			 \frac{\partial M^{nm}_{j}}{\partial Q_z} \mel{k}{Q_z}{j} M^{mg}_{0,k} \langle j | g \rangle
			+M^{nm}_{0,j} \langle k | j \rangle \frac{\partial M^{mg}_{k}}{\partial Q_z} \mel{j}{Q_z}{g} \rbrack \\
		\end{split}
	\end{equation}
	\begin{equation}
	\begin{split}
			B_{2_{ijk}} = \sum_{m,n,j,k,z} \frac{\partial M^{gn}_{i}}{\partial Q_z} \mel{v}{Q_z}{k} M^{nm}_{0,j} 
			M^{mg}_{0,k}
			\langle k | j \rangle
			\langle j | g \rangle
	\end{split}
	\end{equation}
\end{subequations}
where terms such as $\langle a | b \rangle$ are Franck-Condon factors and $\mel{a}{Q_z}{b}$ are Herzberg-Teller coupling factors. %include parameterization of the states?
When $\omega_2$ and $\omega_3$ are nonresonant, the resonance denominators lose their dependence on the vibrational quantum number and the sum can be contracted over ${j, k}$. 
Assuming a normal mode basis, this takes $\sum_{j,k}\langle v | k \rangle \langle k | j \rangle \langle j | g \rangle = \delta_{vg} = 0$, or, A$_{ijk}$ vanishes for singly resonant HDFG. \cite{Neddersen1989}
The B terms do not suffer from this problem.
The Herzberg-Teller matrix elements, where $Q_z$ is a normal mode of the system which is not $\ket{g}$ or $\ket{v}$, allows for coupling between different terms that result in non-zero B terms.
As a result, singly resonant HDFG has significant dependence upon those modes which participate in Herzberg-Teller coupling. 
This likely explains the relative weakness of HDFG in solvents such as hexane. \cite{RN351}
The singly resonant HDFG response can then be approximated as
\begin{equation}
	\gamma_{ijkl} \sim B_{ijk} \mu_l
\end{equation}
or, while a vibrational mode must be infrared active for HDFG output, it must also significantly participate in vibronic coupling, i.e., large B values are a necessity for singly resonant HDFG response.
\subsection{Doubly Resonant}
While HDFG can provide useful information about the properties of molecular vibrations, it is apparent that HDFG can inform on molecular electronic structure by making the hyper-Raman transition resonant with a vibronic state.
DR-HDFG (where a vibrational and vibronic state are resonantly coupled) can provide a tool for measuring vibronic coupling.
This method is analogous to doubly resonant sum frequency generation (DR-SFG). \cite{Shen94}
Many multi-resonant nonlinear spectroscopies have been developed to resolve vibronic coupling in molecular samples. \cite{Carlson1990, Gaynor2017, RN276}
DR-HDFG provides methods to investigate vibronic coupling with only two laser pulses.
Two beam, parametric and non-parametric DR-HDFG processes are diagrammed in \autoref{fig:doubsive}.

To explicitly incorporate vibronic coupling into the selection reules for DR-HDFG, we consider

%todo: write hamiltonian like Gaynor to make clear what we are doing (and put in a picture in Figure doubbleresspec)
To interpret HDFG selection rules, we use a slightly modified version of equation xyz to account for the added vibronic resonance. 
This change immediately makes the A term nonzero in the case of noncentrosymmetric systems.

@@

\subsection{Quantification}

Since the hyper-Raman process is stimulated by a two-photon process, a setup using only two lasers can perform a coherent four wave mixing hyper-Raman experiment by investigating $\pm \vec{k}_1 + 2\vec{k}_2$ processes, assuming a two photon absorption from a single pulse stimulates the hyper-Raman process.
The other commonly used two laser CMDS method is sum frequency generation, a $\chi^{(2)}$ technique, whose phasematching is dictated by $\vec{k}_1 + \vec{k}_2$.
The SFG output scales as $\chi^{(2)}_{IJK} \sim \langle \alpha_{ij} \mu_k \rangle$, where $\alpha_{ij}$ is the Raman polarizability tensor.
Since SFG is dependent upon transitions that transform like both rank-one and rank-two tensors, this enforces a microscopic constraint that SFG cannot yield output for centrosymmetric species under the electric dipole approximation.
This follows from the presence of an inversion center in centrosymmetric systems, which demands strict parity selection rules. \cite{RN230}
Macroscopically, inversion symmetry in the output polarization eliminates output from centrosymmetric species under the electric dipole approximation.\cite{RN227, RN132}
As a result, the output of SFG is significantly reduced relative to most third order spectroscopies because the output only depends on surface species.
However, since HDFG is dependent upon a hyper-Raman transition, it is useful to compare the relative output of the two processes to identify how HDFG output compares to SFG.

@@

With the selection rules of HDFG understood for vibrational spectroscopy, it becomes possible to extract quantitative information from its spectra.
A useful attribute of the $\beta_{ijk}$ dependence in HDFG output is added polarization control. 
Similar to SFG experiments that probe different elements of the infrared transition dipole and Raman polarizability tensors, probing different elements of $\beta_{ijk}$ in HDFG can greatly assist in understanding ultrafast dynamics in isotropic systems, as suggested by Seliya et al. \cite{Shen90, RN224, Bonn2024}
Polarization control in IR-pump-HDFG-probe, analogous to ultrafast transient absorption spectroscopy for vibrational modes, would be an interesting venue to assess the sensitivity of HDFG to different hyper-Raman tensor elements.
